Measurement Bias from Address Aliasing

### 2014-05-30 paper.tex
Add new reference and improvements

% Acknowledgement of support, author(s) address(es)
\begin{bottomstuff}
New Paper, Not an Extension of a Conference Paper.
Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: runeerle@idi.ntnu.no
%LARS: din addresse er ok (NTNU)
\end{bottomstuff}

@@

%\footnote{New Paper, Not an Extension of a Conference Paper.}

@@

The \texttt{cycles_ldm_pending} and \texttt{stalls_l2_pending} are interesting as they are related to the memory and cache system.
However, this program use only 2 cache-lines except for the initial startup.
Using the L1 hit and miss counts also revealed that the L1 load is 99.3\% and store hit rate is ~100\%.
Thus, these events do not indicate a cache hit rate issue, and is probably linked to how the CPU handles address alias.

@@

Because the stack is aligned to 16 bytes, or 4 \emph{words}, there are a couple of different scenarios that could have been observed here.
Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.

A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots. 
While this will give significantly more alias counts, we found that it had little effect on the total number of cycles executed.

In conclusion, we identified that address aliasing is the root cause of measurement bias from environment size for this program.
Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between stack and static data.
The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.

% TODO: Lars: make this a separate section or move it up a bit?
This type of bias can occur in a number of memory configurations.
Here the aliasing was caused by interactions between the stack and static variables.
The stack and heap can interact the same way, depending on how the malloc function behaves.
Since the stack is used for local variables complex code is likely contain many potential conflicts.
A evaluation of how malloc behaves is presented in \ref{sec:heap}.

@@

More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we show the subset that correlates best with the cycle count.
The numbers are for the level 2 optimization case, which we choose to study in more detail.
 % which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:
\begin{itemize}
    \item A high number of resource stalls for the default alignment, which is reduced substantially together with increasing offsets. 
    \item A high number of cycles with memory loads pending, indicating that the pipeline is stalled waiting for load operations to resolve.
    \item Changes to the number of micro-ops executed for certain ports.
\end{itemize}
Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case. 
On Haswell, this port handles various ALU operations, and branching together with port 6.~\cite[Figure 2.1]{OptimizationManual}
As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branch micro-ops are being re-issued.
Unlike the microkernel in \ref{sec:microkernel} the execution and RS stall counts drop with together with aliasing events.
Speculation is possible in this kernel as there are many independent store addresses.
False aliasing limits this speculation however, and makes the CPU incorrectly discard executed instructions.
This leads to the inverted behavior compared to the microkernel.

It is worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
Again the \texttt{cycles_ldm_pending} and \texttt{stalls_l2_pending} behavior is the same as in the microkernel in \ref{sec:microkernel}.
The L1 hit rate remains stable across all offsets, and only a negligible number of memory loads actually misses L1.
On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
These events count the number of outstanding loads to memory outside the processor core each cycle.
Since we do not see any significant number of loads missing L1, the correlation to outstanding loads is probably a result of stalling and more cycles executed.

Overall it seems reasonable to conclude that the resource stalling is causing delayed execution, ultimately generated by false dependencies from address aliasing.
Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
We do not attempt to pinpoint exactly which assembly instructions might alias, but we assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
This is a speedup on top of already aggressive compiler optimization.

@@

Techniques for ``Blind'' optimization using search over variant spaces, is introduced by \citeN{Knights:2009:BlindOpt}.
Their space of variants consist of different position and alignments for each function and global variable, in principle exploring the different configurations possible by altering link order.
The authors of MAO present a framework for extending compilers to provide very low level microarchitectural assembly optimization~\cite{Hundt:2011:MAO}.
Different techniques are used, including rules, pattern matching, and random insertion of nop-instructions.

>>>>

% Acknowledgement of support, author(s) address(es)
\begin{bottomstuff}
New Paper, Not an Extension of a Conference Paper.
Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: larskirk@alumni.ntnu.no, runeerle@idi.ntnu.no
\end{bottomstuff}

@@

@@

The cycles\_ldm\_pending and stalls\_l2\_pending are interesting, as they are related to the memory and cache system.
However, this program use only 2 cache-lines, except for the initial startup.
Using the L1 hit and miss counts also revealed that the L1 load is 99.3\% and store hit rate is ~100\%.
Thus, these events do not indicate a cache hit rate issue, and is probably linked to how the CPU handles address alias.

@@

Because the stack is aligned to 16 bytes, or 4 \emph{words}, there are a couple of different scenarios that could have been observed here.
Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots. 
While this will give significantly more alias counts, we found that it had little effect on the total number of cycles executed.

In conclusion, we identified that address aliasing is the root cause of measurement bias from environment size for this program.
Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between stack and static data.
The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.
This type of bias can occur in a number of memory configurations.
Here, the aliasing was caused by interactions between the stack and static variables, but we can imagine similar conflicts between stack and heap, depending also on how dynamic memory allocators behave.
Because the stack is used for local variables, complex code is likely to contain many potential conflicts with other areas of memory.

@@

More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we show the subset that correlates best with the cycle count.
The numbers are for the level 2 optimization case, which we choose to study in more detail.
However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:
\begin{itemize}
    \item A high number of resource stalls for the default alignment, which is reduced substantially together with increasing offsets. 
    \item A high number of cycles with memory loads pending, indicating that the pipeline is stalled waiting for load operations to resolve.
    \item Changes to the number of micro-ops executed for certain ports.
\end{itemize}
Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case.
On Haswell, this port handles various ALU operations, and branching together with port 6~\cite[Figure 2.1]{OptimizationManual}.
As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branches are being re-issued.
Unlike the microkernel in section \ref{sec:microkernel}, the micro-ops executed and RS stall counts drop together with aliasing events.
Speculation is possible in this kernel, as there are many independent store addresses.
False dependencies will limit the amount of speculation, making the CPU incorrectly discard executed instructions.
This can explain the inverted behavior compared to the microkernel, where the store addresses were fixed between loop iterations.

It is worth noticing that most cache related metrics do \emph{not} stand out in this experiment.
For cycles\_ldm\_pending and stalls\_l2\_pending, we see similar behavior to the microkernel analysis in section \ref{sec:microkernel}.
These events can be explained by pipeline stalls waiting for aliasing store operations to be retired.
The L1 hit rate remains stable across all offsets, and only a negligible number of memory accesses actually miss L1.
On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
These events count the number of outstanding loads to memory outside the processor core each cycle.
Since we do not see any significant number of L1 misses, the correlation to outstanding loads is probably a result of stalling and more cycles executed.

Overall, it seems reasonable to conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
We do not attempt to pinpoint exactly which assembly instructions might alias, but one can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
This is a speedup on top of already aggressive compiler optimization.

@@

Many different approaches have been explored in order to take performance counter data and processor intrinsics into account for optimization.
\citeN{Knights:2009:BlindOpt} introduce ``blind'' optimization, treating the underlying hardware as a black box and use search over \emph{variant spaces}.
Their space of variants consist of different position and alignment for each function and global variable, in principle exploring the different configurations possible by altering link order.
Others have used machine learning to classify performance counter data, recognizing patterns or \emph{pathologies} in measurements to recognize optimization opportunities~\cite{Yoo:ADD}.
The authors of MAO present a framework for extending compilers to provide very low level microarchitectural assembly optimization, using rules, pattern matching, and random insertion of nop-instructions to discover optimal assembly outputs~\cite{Hundt:2011:MAO}.
